\mytitle{}

### 2019-01-11 paper/paper.tex
Appease the PCs several months in advance.
WIP on explanations/justifications

% !TeX document-id = {e7fef4b9-afbc-4e81-94aa-c4847183d047}
% !BIB program = biber
\documentclass[conference, letterpaper, 10pt, times]{IEEEtran}
\usepackage{nopageno}

@@

\usepackage{url}
\usepackage{hyperref}
\usepackage{cleveref}
\newcommand{\crefrangeconjunction}{--}

\hypersetup{
	colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=black,
	pdftitle={\mytitle{}},
%	pdfauthor={Kyle A. Simpson, Dimitrios P. Pezaros, Simon Rogers}
	pdfauthor={Anonymous Giraffe, Anonymous Badger, Anonymous Ibex}
}
\newcommand*{\email}[1]{\href{mailto:#1}{\nolinkurl{#1}} }

@@

@@

\title{\mytitle{}}
\author{
%	Kyle A.\ Simpson\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant number EP/M508056/1].},
%	Kyle A.\ Simpson,
%	Dimitrios P.\ Pezaros,
%   Simon Rogers\\
%	University of Glasgow, Glasgow, Scotland,\\
%	\email{k.simpson.1@research.gla.ac.uk}
Anonymous Giraffe,
Anonymous Badger,
Anonymous Ibex\\
Unnamed Department, Nowhere\\
\email{giraffe.a@unnamed.com}
}

@@

\emph{Reinforcement learning} (RL) may overcome the detection problem for certain classes of anomaly by managing and monitoring \emph{consequences}; an agent's role is to learn to optimise performance criteria (which are always available).
We examine how the paradigm has been applied to the task of DDoS prevention to date, and present an analysis of suspected edge cases to uncover existing weaknesses in present approaches.
These flaws---poor handling of TCP traffic and high host density at egress points---inspire and direct an exploration of the design space of RL models.
We advance the state-of-the-art of DDoS mitigation using RL by introducing two agent classes designed to act on a per-flow basis, in a protocol-agnostic manner.
This is supported by an in-depth investigation of feature suitability and empirical evaluation.

@@

While \emph{machine learning} (ML) approaches seem like a sensible fit for this problem, in \citeyear{DBLP:conf/sp/SommerP10} \citeauthor{DBLP:conf/sp/SommerP10} identified the `failure to launch' of ML-based anomaly detection systems, and to quite a large extent this remains the case today \cite{DBLP:conf/sp/SommerP10}.
Their application is made difficult due to significant operational differences from standard ML tasks, alongside certain characteristics of network traffic.
Briefly, these are the diversity of network traffic across varying timescales \cite{DBLP:conf/sp/SommerP10} and significant burstiness \cite{DBLP:journals/ccr/LelandWTW95}.
Above the aggregate level, the constant deployment of new services and new protocols means that traffic is \emph{non-stationary} and displays an evolving notion of normality.
Further issues arise from the extraordinarily low tolerance for false positives inherent to network intrusion detection \cite{DBLP:conf/ccs/Axelsson99}, and the challenges encountered when learning from unlabelled (often partial) data.
All of these factors greatly inflate the difficulty of the detection problem.

For certain classes of problem e.g., flooding-based DDoS attacks, \emph{reinforcement learning} (RL) offers another perspective.
The role of a RL agent differs from that of a standard classifier, adaptively reacting to threats by assuming the role of a feedback loop for network optimisation, typically to safeguard service guarantees.
In a sense, this allows us to ``overcome'' the difficulties of the detection problem by monitoring \emph{performance characteristics and consequences} in real-time; by looking for the effect rather than the cause.
The intent is to augment what existing misuse-based solutions can provide, by automatically alerting, recording and controlling what are believed to be illegal system states.
%Whether it takes direct control of the network, or is used indirectly to optimise a key part of another system, more powerful `deep' RL techniques (and well-founded action spaces) aren't yet well explored for network IDS/IPS.
%These range from more modern training algorithms \cite{DBLP:journals/corr/SchulmanWDRK17, DBLP:conf/icml/SchulmanLAJM15}, to evolutionary strategies \cite{DBLP:journals/corr/SalimansHCS17, DBLP:journals/corr/abs-1802-08842}, hierarchical action composition \cite{DBLP:journals/corr/abs-1710-09767}, and competitive multi-agent learning \cite{DBLP:journals/corr/abs-1710-03748}.

@@

%?? Discuss the most important conclusions before the outline.
We find that the existing work for this task fails to account for congestion-aware traffic (i.e., TCP) and deployment environments with high host density per egress point, achieving poor results due to an overly coarse view of the network.
The approach we identify to remedy this is to make throttling decisions on a per-source basis, and we present the engineering decisions this mandates: updating RL agents from multiple traces per timestep, timed random sequential action computation and a supporting SDN architecture.
In tandem with the development and evaluation of an effective state-space and model, we provide the design of a second model inspired by past work on algorithmic DDoS prevention, as an example of the integration of domain-specific knowledge.
Our introduction of per-source decisions improves substantially upon the state-of-the-art, and we show that our second model achieves excellent performance for large $n$.
Crucially, both models remain protocol- and content-agnostic to offer future-proofing against the rollout of future protocols like QUIC \cite{DBLP:conf/sigcomm/LangleyRWVKZYKS17}.
%?? Also algorithmic enhancements such as multiple actions per timestep, 
%?? PROTOCOL-AGNOSTIC -- HOW WILL THESE THINGS COPE WITH QUIC ET AL.?!?!
Briefly, this paper contributes:
\begin{itemize}
%	\item A DDoS mitigation system based on direct-control reinforcement learning, designed for deployment in real-world software-defined networks (\cref{sec:environment-and-rl-algorithm,sec:acting-on-individual-sources}).
	\item Important weaknesses and flaws in the past design and evaluation of similar techniques \cite{DBLP:journals/eaai/MalialisK15}. We offer a deconstruction and empirical study on their formulation, flaws, and risk factors concerning different traffic distributions---in addition, we demonstrate modifications enabling deployment in real-world software-defined networks (\cref{sec:performance-in-an-emulated-environment}).
	\item A reactive simulation of web-server traffic, designed to test system characteristics that packet trace playback fails to capture (\cref{sec:a-new-normal}).
	\item An in-depth investigation into suitable features for an automatic DDoS mitigation system, with qualitative and quantitative justification (\cref{sec:rethinking-the-state-space}).
	\item Two source-level granularity approaches to RL-driven DDoS prevention (\emph{Marl++}, \emph{SPF}), improving upon past aggregate-based models (\cref{sec:acting-on-individual-sources}), alongside an empirical evaluation (\cref{sec:the-results-of-doing-so}) and discussion of security concerns and real-world deployment (\cref{sec:discussion}).
\end{itemize}

\section{Background and Threat Model}
%?? Introduce RL, related definitions etc.
\subsection{Reinforcement Learning}\label{sec:reinforcement-learning}
\emph{Reinforcement learning} (RL) is a variant of machine learning principally concerned with training an agent to choose an optimal sequence of actions in the pursuit of a given task \cite{RL2E}.
We assume the agent has a certain amount of knowledge whenever a decision must be made: at any point in time $t$, it knows which \emph{state} it is in ($S_t \in \mathcal{S}$), the set of \emph{actions} which are available to it ($\operatorname{A}(S_t) \subseteq \mathcal{A}$) and, if available, a numeric \emph{reward} obtained from the last action chosen ($R_t \in \mathbb{R}, A_{t-1} \in \operatorname{A}(S_{t-1})$).
This model of system interaction is best described as a \emph{Markov Decision Process} (MDP).
RL methods combine this information with a current \emph{policy} $\pi$ to determine which action should be taken: such a choice need not be optimal if an agent needs to further explore some region of the state space.
The policy is then further refined by updating value estimates for state-action pairs or via policy gradient methods, meaning that RL-based approaches learn adaptively and online if reward functions are available in the environment they are deployed in.
In practice, this means that agents are able to automatically adapt to evolving problems without operator intervention or a new, custom-built training corpus.

@@

There is immense variation in \emph{how} policies and/or values may be learned, reliant upon the learning environment, problem and required convergence guarantees.
In particular, we shall focus on methods which choose actions according to their value estimates from the current state.
To cope with a continuous state- and/or action-space, one valuable technique is to employ \emph{tile-coding} \cite[pp.\ \numrange{217}{221}]{RL2E}---this converts our state from $\mathbb{R}^d \times \mathcal{A}$ into a sparse boolean feature vector $\operatorname{\mathbf{x}}(s, a)$ by subdividing it into a number of overlapping $(d+ \dim{\mathcal{A}})$-dimensional grids with an optional bias component.
The numbers of tilings and of tiles per dimension have a direct effect on the resolution of the discretisation.
Entries of $\operatorname{\mathbf{x}}(s, a)$ are set to 1 if the state-action pair exists within the corresponding tile.
This transform, as an example of linear function approximation, then necessitates the use of \emph{1-step semi-gradient Sarsa} \cite[pp.\ \numrange{243}{244}]{RL2E}.
Given a weight vector $\wvec{0}=\bm{0}$ and learning rate $\alpha$, we may approximate an action $a$'s value in any state $s$ taking $\operatorname{q}(s, a) \approx \acval{s}{a}{\wvec{}}$ by continually updating $\wvec{t}$ as follows:
\begin{subequations}
	\begin{gather}
	\acval{s}{a}{\wvec{}} = \wvec{}^{\top} \operatorname{\mathbf{x}}(s, a),\\
	\delta_t = R_{t+1} + \gamma \acval{S_{t+1}}{A_{t+1}}{\wvec{t}} - \acval{S_t}{A_t}{\wvec{t}},\\
	\bm{w}_{t+1} = \bm{w}_{t} + \alpha \delta_t \nabla{\acval{S_t}{A_t}{\wvec{t}}},
	\end{gather}
	\label{eqn:sg-sarsa}
	taking $\nabla$ with respect to $\wvec{}$.
\end{subequations}

Computing the approximate value of every action available in the current state forms the basis of a policy.
Actions with maximal value can be chosen each time (the \emph{greedy} policy), we might modify this by taking random actions with probability $\epsilon$ to encourage early exploration (the \emph{$\epsilon$-greedy policy}), or we might use some other mechanism.

This combination of algorithm and coding strategy is well-optimised.
If $\operatorname{A}(s) = \mathcal{A}$ in every state (and actions are themselves discrete), these properties allow a particularly efficient (vectorised) implementation of the policy and update rules by storing a vector of action values for each state encountered.
Action values for any state are then obtained by summation of the weight vectors from all activated tiles---taking $|\mathcal{A}|(n_{\mathit{tilings}}-1)$ floating point additions per decision.
Observing that $\nabla{\acval{s}{a}{\wvec{}}} = \operatorname{\mathbf{x}}(s, a)$, further optimisations arise by considering that a tile-coded feature vector is a binary vector of constant Hamming weight (and so is amenable to representation as an array of indices, $s_{\mathit{list}}$).
This means that we need only perform $n_{\mathit{tilings}} + 2$ additions and \num{2} multiplications per model update:
\begin{equation}
	\bm{w}_{t+1}[i][\operatorname{index}(A_t)] = \bm{w}_{t}[i][\operatorname{index}(A_t)] + \alpha \delta_t, \forall i \in s_{\mathit{list}}.
	\label{eqn:sg-sarsa-opt}
\end{equation}
If desired we may define a state-space with an arbitrary number of tiles per dimension (higher-resolution, lower generalisation), yet having constant-size state vectors and constant action computation cost ($\mathcal{O}(n_{\mathit{tilings}})$).
Beyond this, we need not store action values for tiles which have not yet been visited, conserving memory.

@@

\subsection{Distributed Denial of Service (DDoS)}
%?? DDoS attack variants (leading into characteristics, supporting features). Amplification (UDP \cite{DBLP:conf/ndss/Rossow14}, TCP \cite{DBLP:conf/uss/KuhrerHRH14}), Transit-link (Crossfire \cite{DBLP:conf/sp/KangLG13}, Coremelt \cite{DBLP:conf/esorics/StuderP09}). Mirai botnet's involvement \cite{DBLP:conf/uss/AntonakakisABBB17}.

>>>>

% !TeX document-id = {e7fef4b9-afbc-4e81-94aa-c4847183d047}
% !BIB program = biber
\documentclass[conference, a4paper, 10pt, times]{IEEEtran}
\usepackage{nopageno}

@@

\usepackage{url}
\usepackage{hyperref}
\usepackage[nameinlink]{cleveref}
\newcommand{\crefrangeconjunction}{--}

\hypersetup{
	colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=black,
	pdftitle={\mytitle{}},
%	pdfauthor={Kyle A. Simpson, Simon Rogers, Dimitrios P. Pezaros}
	pdfauthor={Anonymous Giraffe, Anonymous Badger, Anonymous Ibex}
}
\newcommand*{\email}[1]{\href{mailto:#1}{\nolinkurl{#1}} }

@@

\definecolor{inferno0}{rgb}{0.001462 0.000466 0.013866}
\definecolor{inferno64}{rgb}{0.341500 0.062325 0.429425}
\definecolor{inferno128}{rgb}{0.735683 0.215906 0.330245}
\definecolor{inferno192}{rgb}{0.978422 0.557937 0.034931}
\definecolor{inferno255}{rgb}{0.988362 0.998364 0.644924}

\colorlet{lowac}{inferno128}
\colorlet{midac}{inferno192}
\colorlet{highac}{inferno255}

@@

\title{\mytitle{}}
\author{
%	Kyle A.\ Simpson\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant number EP/M508056/1].},
	Kyle A.\ Simpson,
	Simon Rogers,
	Dimitrios P.\ Pezaros,\\
	University of Glasgow, Glasgow, Scotland,\\
	\email{k.simpson.1@research.gla.ac.uk}
%Anonymous Giraffe,
%Anonymous Badger,
%Anonymous Ibex\\
%Unnamed Department, Nowhere\\
%\email{giraffe.a@unnamed.com}
}

@@

\emph{Reinforcement learning} (RL) may overcome the detection problem for certain classes of anomaly by managing and monitoring \emph{consequences}; an agent's role is to learn to optimise performance criteria (which are always available).
We examine how the paradigm has been applied to the task of \emph{Distributed Denial of Service} (DDoS) attack prevention to date, and present an analysis of suspected edge cases to uncover existing weaknesses in present approaches.
These flaws---poor handling of TCP traffic and high host density at egress points---inspire and direct an exploration of the design space of RL models.
We advance the state-of-the-art of DDoS mitigation using RL by introducing two agent classes designed to act on a per-flow basis, in a protocol-agnostic manner.
This is supported by an in-depth investigation of feature suitability and empirical evaluation.

@@

While \emph{machine learning} (ML) approaches seem like a sensible fit for this problem, in \citeyear{DBLP:conf/sp/SommerP10} \citeauthor{DBLP:conf/sp/SommerP10} identified the `failure to launch' of ML-based anomaly detection systems---a distinct lack of real-world system deployments \cite{DBLP:conf/sp/SommerP10}.
To quite a large extent, this remains the case today.
?? RESTRUCTURE
These authors posit that their application is made difficult due to significant operational differences from standard ML tasks.
?? (quoted: high cost of errors, lack of training data, semantic gap between results and operational interpretation, variability in input data, fundamental difficulties for conducting sound evaluation)
This is further aggravated by certain characteristics of network traffic; the diversity of network traffic across varying timescales \cite{DBLP:conf/sp/SommerP10} and significant burstiness \cite{DBLP:journals/ccr/LelandWTW95}.
Above the aggregate level, the constant deployment of new services and new protocols means that traffic is \emph{non-stationary} and displays an evolving notion of normality.
Further issues arise from the extraordinarily low tolerance for false positives inherent to network intrusion detection \cite{DBLP:conf/ccs/Axelsson99}, and the challenges encountered when learning from unlabelled (often partial) data.
All of these factors greatly inflate the difficulty of the detection problem.

?? Make it clearer here what problem I specifically want to solve: principally a particular class of DDoS attacks; volume-based DDoS attacks. Amplification attacks are just a specialisation, this can be made more obvious. I think I need to be clearer about the \emph{intended deployment environment} of service hosts (i.e., not ISPs).

For certain classes of problem e.g., flooding-based DDoS attacks, \emph{reinforcement learning} (RL) offers another perspective.
?? Unclear explanation of RL here?
The role of a RL agent differs from that of a standard classifier, adaptively reacting to threats by assuming the role of a feedback loop for network optimisation, typically to safeguard service guarantees.
In a sense, this allows us to ``overcome'' the difficulties of the detection problem by monitoring \emph{performance characteristics and consequences} in real-time; by looking for the effect rather than the cause.
The intent is to augment what existing misuse-based solutions can provide, by automatically alerting, recording and controlling what are believed to be illegal system states.
%Whether it takes direct control of the network, or is used indirectly to optimise a key part of another system, more powerful `deep' RL techniques (and well-founded action spaces) aren't yet well explored for network IDS/IPS.
%These range from more modern training algorithms \cite{DBLP:journals/corr/SchulmanWDRK17, DBLP:conf/icml/SchulmanLAJM15}, to evolutionary strategies \cite{DBLP:journals/corr/SalimansHCS17, DBLP:journals/corr/abs-1802-08842}, hierarchical action composition \cite{DBLP:journals/corr/abs-1710-09767}, and competitive multi-agent learning \cite{DBLP:journals/corr/abs-1710-03748}.

@@

%?? Discuss the most important conclusions before the outline.
We find that the existing work for this task fails to account for congestion-aware traffic (i.e., TCP) and deployment environments with high host density per egress point, achieving poor results due to an overly coarse view of the network.
The approach we identify to remedy this is to make throttling decisions on a per-source basis, and we present the engineering decisions this mandates: updating RL agents from multiple traces per timestep, timed random sequential action computation and a supporting SDN architecture.
In tandem with the development and evaluation of an effective state-space and model, we provide the design of a second model inspired by past work on algorithmic DDoS prevention, as an example of the integration of domain-specific knowledge.
Our introduction of per-source decisions improves substantially upon the state-of-the-art, and we show that our second model achieves excellent performance for high host density.
Crucially, both models remain protocol- and content-agnostic to offer future-proofing against the rollout of future protocols like QUIC \cite{DBLP:conf/sigcomm/LangleyRWVKZYKS17}.
%?? Also algorithmic enhancements such as multiple actions per timestep, 
%?? PROTOCOL-AGNOSTIC -- HOW WILL THESE THINGS COPE WITH QUIC ET AL.?!?!
Briefly, this paper contributes:
\begin{itemize}
%	\item A DDoS mitigation system based on direct-control reinforcement learning, designed for deployment in real-world software-defined networks (\cref{sec:environment-and-rl-algorithm,sec:acting-on-individual-sources}).
	\item Important weaknesses and flaws in the past design and evaluation of similar techniques \cite{DBLP:journals/eaai/MalialisK15}. We offer a deconstruction and empirical study on their formulation, flaws, and risk factors concerning different traffic distributions---in addition, we demonstrate modifications enabling deployment in real-world software-defined networks (\cref{sec:performance-in-an-emulated-environment}).
	\item A reactive simulation of web-server traffic, designed to test system characteristics that packet trace playback fails to capture (\cref{sec:a-new-normal}).
	\item An in-depth investigation into suitable features for an automatic DDoS mitigation system, with qualitative and quantitative justification (\cref{sec:rethinking-the-state-space}).
	\item Two source-level granularity approaches to RL-driven DDoS prevention (\emph{Marl++}, \emph{SPF}), improving upon past aggregate-based models (\cref{sec:acting-on-individual-sources}), alongside an empirical evaluation (\cref{sec:the-results-of-doing-so}) and discussion of security concerns and real-world deployment (\cref{sec:discussion}).
\end{itemize}

\section{Background and Threat Model}
%?? Introduce RL, related definitions etc.

\subsection{Distributed Denial of Service}
%?? DDoS attack variants (leading into characteristics, supporting features). Amplification (UDP \cite{DBLP:conf/ndss/Rossow14}, TCP \cite{DBLP:conf/uss/KuhrerHRH14}), Transit-link (Crossfire \cite{DBLP:conf/sp/KangLG13}, Coremelt \cite{DBLP:conf/esorics/StuderP09}). Mirai botnet's involvement \cite{DBLP:conf/uss/AntonakakisABBB17}.

\emph{Distributed denial of service} (DDoS) attacks are concentrated efforts by many hosts to reduce the availability of a service, typically to inflict financial harm or as an act of vandalism.
Attackers achieve this either by exploiting peculiarities of operating system or application behaviour (e.g., \emph{SYN flooding attacks}), or overwhelming their target through sheer volume of requests or inbound packets (\emph{flooding-based attacks}).
Hosts often participate unwillingly, typically having been recruited into a \emph{botnet} by malware infection to be orchestrated from elsewhere \cite{DBLP:conf/uss/AntonakakisABBB17}.

?? EXPAND

\subsection{Reinforcement Learning}\label{sec:reinforcement-learning}
\emph{Reinforcement learning} (RL) is a variant of machine learning principally concerned with training an agent to choose an optimal sequence of actions in the pursuit of a given task \cite{RL2E}.
We assume the agent has a certain amount of knowledge whenever a decision must be made: at any point in time $t$, it knows which \emph{state} it is in ($S_t \in \mathcal{S}$), the set of \emph{actions} which are available to it ($\operatorname{A}(S_t) \subseteq \mathcal{A}$) and, if available, a numeric \emph{reward} obtained from the last action chosen ($R_t \in \mathbb{R}, A_{t-1} \in \operatorname{A}(S_{t-1})$).
This model of system interaction is best described as a \emph{Markov Decision Process} (MDP).
RL methods combine this information with a current \emph{policy} $\pi$ to determine which action should be taken: such a choice need not be optimal if an agent needs to further explore some region of the state space.
The policy is then further refined by updating value estimates for state-action pairs or via policy gradient methods, meaning that RL-based approaches learn adaptively and online if reward functions are available in the environment they are deployed in.
In practice, this means that agents are able to automatically adapt to evolving problems without operator intervention or a new, custom-built training corpus.

@@

There is immense variation in \emph{how} policies and/or values may be learned, reliant upon the learning environment, problem and required convergence guarantees.
In particular, we shall focus on methods which choose actions according to their value estimates from the current state: let $\operatorname{q}(s, a) \in \mathbb{R}$ denote the estimate of action $a$'s value if it were to be taken in state $s$.
%?? Revisit the maths here, d, dim A, s...
Exact (tabular) representations require that we store a value estimate for each action in every state---if state is real-valued or high-dimensional, then computation and storage quickly become infeasible.
To cope with a continuous state and/or action space, one valuable technique is to employ linear function approximation backed by \emph{tile coding} \cite[pp.\ \numrange{217}{221}]{RL2E}.

\begin{figure}
	\centering
	\begin{subfigure}{0.4\linewidth}
		\resizebox{\linewidth}{!}{
		\begin{tikzpicture}
		\node at (0,0.3){$s = \begin{pmatrix}
			0.7 \\
			0.3
			\end{pmatrix}$};
		\node at (0,-1) {$\bm{x}(s,\cdot) = \begin{Bmatrix}
			T_{1,9}, \\
			T_{2,5}, \\
			T_{\mathit{bias}}
			\end{Bmatrix}$};

		\node at (2.5,-0.5) {
			\begin{tikzpicture}
			\draw[step=0.5cm,color=uofgcobalt,opacity=0.7,shift={(0,0)},label=above:{Tiling 0}] (-0.5,-0.5) grid (1,1);
			\fill[uofgcobalt,opacity=0.5] (0.5,-0.5) rectangle (1,0);
			\node[color=uofgcobalt] at (0,1.1) {\footnotesize Tiling 1};

			\draw[step=0.5cm,color=uofgpumpkin,opacity=0.7,shift={(0.25,-0.25)},label=above:{Tiling 1}] (-0.5,-0.5) grid (1,1);
			\fill[uofgpumpkin,opacity=0.5,shift={(0.25,-0.25)}] (0,0) rectangle (0.5,0.5);
			\node[color=uofgpumpkin] at (0.25,-0.95) {\footnotesize Tiling 2};

			\node[circle, black, draw,
			fill, radius=0.5pt, inner sep=0pt,minimum size=1.5pt, label=above:{$s$}] at (0.625,-0.125) {};
%			\filldraw (0.625,-0.125) circle[radius=1.5pt,label=above:{$s$}];

			\draw[->] (-0.25,-0.5)--(-0.25,0.85);
			\draw[->] (-0.25,-0.5)--(1.1,-0.5);

			\node at (1,-0.7) {\footnotesize 1};
			\node at (-0.4,0.75) {\footnotesize 1};
			\node at (-0.35,-0.6) {\footnotesize 0};
			\end{tikzpicture}
		};

		\end{tikzpicture}
		}
		\caption{Tile Coding\label{fig:tilecode-code}}
	\end{subfigure}%
	\begin{subfigure}{0.4\linewidth}
		\resizebox{\linewidth}{!}{
		\begin{tikzpicture}
		% Top half
		\def\topacs{
			{-0.3,-0.5,-0.1,0.8},
			{0.1,0.1,-0.2,0.3},
			{0.3,0.4,0.2,-0.4}%
		}

		\foreach \line [count=\y] in \topacs {
			\foreach \pix [count=\x] in \line {
				\ifthenelse{\lengthtest{\pix pt < 0 pt}}{
					\pgfmathtruncatemacro\lambda{(\pix+1)*100}
					\draw[shift={(-0.7,0)},fill=midac!\lambda!lowac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}{
					\pgfmathtruncatemacro\lambda{\pix*100}
					\draw[shift={(-0.7,0)},fill=highac!\lambda!midac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}
				\node[shift={(-0.35,0.175)}] at (0.7*\x,-0.35*\y) {\footnotesize \pix};
			}
		}

		\draw[xstep=0.7cm,ystep=0.35cm,shift={(0,0)}] (0,-1.06) grid (2.8,0);
		\node[label=left:{$T_{1,9}$},shift={(0,-0.125)}] at (0,0) {}; 
		\node[label=left:{$T_{2,5}$},shift={(0,-0.125)}] at (0,-0.375) {}; 
		\node[label=left:{$T_{\mathit{bias}}$},shift={(0,-0.125)}] at (0,-0.75) {};

		% bottom half
		\def\botacs{
			{0.1,0.0,-0.1,0.7}%
		}

		\foreach \line [count=\y] in \botacs {
			\foreach \pix [count=\x] in \line {
				\ifthenelse{\lengthtest{\pix pt < 0 pt}}{
					\pgfmathtruncatemacro\lambda{(\pix+1)*100}
					\draw[shift={(-0.7,-1.275)},fill=midac!\lambda!lowac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}{
					\pgfmathtruncatemacro\lambda{\pix*100}
					\draw[shift={(-0.7,-1.275)},fill=highac!\lambda!midac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}
				\node[shift={(-0.35,-1.1)}] at (0.7*\x,-0.35*\y) {\footnotesize \pix};
			}
		}

		\draw[xstep=0.7cm,ystep=0.35cm,shift={(0,-1.275)}] (0,-0.35) grid (2.8,0);
		\node[label=left:{$\mathit{Total}$},shift={(0,-0.175)}] at (0,-1.275) {};

		\draw [->] (2.45, -1.9) -- (2.45, -1.65);
		\end{tikzpicture}
		}
		\caption{Action Selection\label{fig:tilecode-select}}
	\end{subfigure}

	\caption{
		An example of tile coding for 2-dimensional state and 4 actions, using 2 tilings, 3 tiles per dimension, and a bias tile.
		All components of $s$ are clamped to the range $[0,1]$.
		For simplicity, we denote $\bm{x}(s,\cdot)$ as a list of indices and represent the values of all actions at each tile with a vector.
		(a) The state $s$ activates the bias tile and exactly one tile in each tiling.
		(b) The action values of active tiles are summed to produce the current value estimate for each action available in $s$---for this state, local knowledge ensures that action 4 is chosen by the greedy policy despite typically being a poor choice elsewhere.
		\label{fig:tilecode}
	}
\end{figure}

Tile coding is a form of feature representation which converts a state-action pair into a sparse boolean feature vector $\operatorname{\mathbf{x}}(s, a)$ by subdividing a $d$-dimensional subset of the space into a number of overlapping grids with an optional bias component.
Each tile corresponds to an entry of $\operatorname{\mathbf{x}}(s, a)$ which is set to 1 if the state-action pair lies within it.
\Cref{fig:tilecode-code} demonstrates the process for a 2-dimensional state space, and how the numbers of tilings and tiles per dimension control feature resolution and generalisation.
Moreover, to capture combinatorial effects or create multi-scale representation we may combine codings by concatenating individual feature vectors.
We may then approximate an action's value with respect to a policy parameter vector $\wvec{}$, defining some $\acval{s}{a}{\wvec{}} \approx \operatorname{q}(s, a)$:
\begin{equation}
%	\begin{gather}
\acval{s}{a}{\wvec{}} = \wvec{}^{\top} \operatorname{\mathbf{x}}(s, a)
%	\end{gather}
\label{eqn:lin-approx}
\end{equation}
As each component of $\wvec{}$ is the value estimate of the corresponding tile, learning an effective policy is equivalent to learning $\wvec{}$.
Given a learning rate $\alpha \in \mathbb{R}$ and initialising $\wvec{0}=\bm{0}$, we may continually update $\wvec{t}$ using the \emph{1-step semi-gradient Sarsa} algorithm \cite[pp.\ \numrange{243}{244}]{RL2E}:
\begin{subequations}
	\begin{gather}
	\delta_t = R_{t+1} + \gamma \acval{S_{t+1}}{A_{t+1}}{\wvec{t}} - \acval{S_t}{A_t}{\wvec{t}},\\
	\bm{w}_{t+1} = \bm{w}_{t} + \alpha \delta_t \nabla{\acval{S_t}{A_t}{\wvec{t}}},
	\end{gather}
	\label{eqn:sg-sarsa}
	where $\delta_t$ is known as the \emph{temporal-difference} (TD) error, and the vector gradient $\nabla$ is taken with respect to $\wvec{}$.
\end{subequations}

Computing the approximate value of every action available in the current state forms the basis of a policy.
Actions with maximal value can be chosen each time (the \emph{greedy} policy), we might modify this by taking random actions with probability $\epsilon$ to encourage early exploration (the \emph{$\epsilon$-greedy policy}), or we might use some other mechanism.
\Cref{fig:tilecode-select} extends our working example to show how the value of each action is computed, combining a global estimate ($T_{\mathit{bias}}$) with knowledge particular to each state.

\begin{figure}
	\centering
	\resizebox{0.5\linewidth}{!}{
		\begin{tikzpicture}		
		% Top half
		\def\startacs{
			{0.8},
			{0.3},
			{-0.4}%
		}

		\foreach \line [count=\y] in \startacs {
			\foreach \pix [count=\x] in \line {
				\ifthenelse{\lengthtest{\pix pt < 0 pt}}{
					\pgfmathtruncatemacro\lambda{(\pix+1)*100}
					\draw[shift={(-0.7,0)},fill=midac!\lambda!lowac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}{
					\pgfmathtruncatemacro\lambda{\pix*100}
					\draw[shift={(-0.7,0)},fill=highac!\lambda!midac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}
				\node[shift={(-0.35,0.175)}] at (0.7*\x,-0.35*\y) {\footnotesize \pix};
			}
		}

		\draw[xstep=0.7cm,ystep=0.35cm,shift={(0,0)}] (0,-1.06) grid (0.7,0);
		\node[label=left:{$T_{1,9}$},shift={(0,-0.125)}] at (0,0) {}; 
		\node[label=left:{$T_{2,5}$},shift={(0,-0.125)}] at (0,-0.375) {}; 
		\node[label=left:{$T_{\mathit{bias}}$},shift={(0,-0.125)}] at (0,-0.75) {};
		\node[label=below:{\footnotesize Action 4},shift={(0.35,-0.125)}] at (0,-0.75) {};
		\node[label=above:{\footnotesize $\wvec{t}$},shift={(0.35,-0.125)}] at (0,0) {};

		\def\finalacs{
			{0.7},
			{0.2},
			{-0.5}%
		}

		\foreach \line [count=\y] in \finalacs {
			\foreach \pix [count=\x] in \line {
				\ifthenelse{\lengthtest{\pix pt < 0 pt}}{
					\pgfmathtruncatemacro\lambda{(\pix+1)*100}
					\draw[shift={(2-0.7,0)},fill=midac!\lambda!lowac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}{
					\pgfmathtruncatemacro\lambda{\pix*100}
					\draw[shift={(2-0.7,0)},fill=highac!\lambda!midac] (0.7*\x,-0.35*\y) rectangle +(0.7,0.35);
				}
				\node[shift={(2-0.35,0.175)}] at (0.7*\x,-0.35*\y) {\footnotesize \pix};
			}
		}

		\draw[xstep=0.7cm,ystep=0.35cm,shift={(2,0)}] (0,-1.06) grid +(0.7,1.06);
		\node[label=below:{\footnotesize Action 4},shift={(2.35,-0.125)}] at (0,-0.75) {};
		\node[label=above:{\footnotesize $\wvec{t+1}$},shift={(0.35,-0.125)}] at (2,0) {};

		\draw[->] (0.9,-0.5) -- node[above] {$+ \alpha \delta_t$} (1.8,-0.5);
		\end{tikzpicture}
	}

\caption{
	The update step for \cref{fig:tilecode}, given an observed TD error $\delta_t=-0.2$ (indicating a lower observed reward than the expected long-term value of \num{0.7}) and $\alpha=0.5$.
	Action 4's value is thus reduced in the tiles associated with state $s$, but remains the most likely choice; the negative $delta_t$ may have arisen from noise in the reward signal.
	For illustrative purposes, we choose an unrealistically high $\alpha$ (typically, $\alpha=0.05$ or lower is a more reasonable choice).
	\label{fig:tilecode-update}
}
\end{figure}

This combination of algorithm and coding strategy is well-optimised.
If $\operatorname{A}(s) = \mathcal{A}$ in every state (and actions are themselves discrete), these properties allow a particularly efficient (vectorised) implementation of the policy and update rules by storing a vector of action values for each state encountered.
Action values for any state are then obtained by summation of the weight vectors from all activated tiles---taking $|\mathcal{A}|(n_{\mathit{tilings}}-1)$ floating point additions per decision.
Observing that $\nabla{\acval{s}{a}{\wvec{}}} = \operatorname{\mathbf{x}}(s, a)$, further optimisations arise by considering that a tile-coded feature vector is a binary vector of constant Hamming weight (and so is amenable to representation as an array of indices, $s_{\mathit{list}}$).
This means that we need only perform $n_{\mathit{tilings}} + 2$ additions and \num{2} multiplications per model update:
\begin{equation}
\bm{w}_{t+1}[i][\operatorname{index}(A_t)] = \bm{w}_{t}[i][\operatorname{index}(A_t)] + \alpha \delta_t, \forall i \in s_{\mathit{list}}.
\label{eqn:sg-sarsa-opt}
\end{equation}
\Cref{fig:tilecode-update} shows how this applies to our prior example.
If desired we may define a state-space with an arbitrary number of tiles per dimension (higher-resolution, lower generalisation), yet having constant-size state vectors and constant action computation cost ($\mathcal{O}(n_{\mathit{tilings}})$).
Beyond this, we need not store action values for tiles which have not yet been visited, conserving memory.

@@
